\textbf{SPECFEM3D Cartesian}\\ \textbf{User Manual}

### 2015-10-21 doc/USER_MANUAL/05_running_the_solver.tex
better version of the notes on the acoustic formulation

In SPECFEM, when considering purely acoustic waves (neglecting gravity acceleration and viscosity)
in fluid regions one uses a potential formulation to represent the "density times displacement" perturbation
\cite{ChKoViCaVaFe07}:
\begin{equation}
\rho \mathbf{u} = \nabla \chi
\label{potential}
\end{equation}
where $\mathbf{u}$ is the displacement perturbation, $\rho$ the background density and $\chi$ is the scalar potential.\\
This formulation has two great advantages: it reduces the number of unknowns from three to one, and it also automatically removes artefacts
in finite-element methods (FEM) caused by the displacement formulation \cite{HaOuVe78}.
This is crucial since it is the main reason for which a displacement formulation cannot be used in fluid regions in a FEM.
\newline

@@

The solution to this is to work with pressure as the scalar unknown (\cite{God11}, equation 23),
or even better with the time integral (the primitive) of pressure (see e.g. \cite{Eve81});
this was the approach that was originally implemented in SPECFEM until 10 years ago or so, but \cite{ChCaVi03} showed that switching to a displacement potential
or to a potential of $\rho \mathbf{u}$ enables one to couple the fluid and solid regions without iterating on the coupling condition in the time scheme,
thus resulting in much easier and cheaper implementation; since at that time SPECFEM users were not studying media with strong local variations of density (in solid Earth
geophysics such large local variations in fluid regions never happen), the SPECFEM developers permanently switched to such a displacement-based potential formulation.
However, for media with strong local variations of density in a fluid region we should reconsider the issue and consider trying the pressure-based formulation again.

>>>>

In SPECFEM, when considering purely acoustic waves (neglecting gravity acceleration and viscosity)
in fluid regions one uses a potential formulation to represent the "density times displacement" perturbation
\citep{ChKoViCaVaFe07}:
\begin{equation}
\rho \mathbf{u} = \nabla \chi
\label{potential}
\end{equation}
where $\mathbf{u}$ is the displacement perturbation, $\rho$ the background density and $\chi$ is the scalar potential.\\
This formulation has several great advantages: it reduces the number of unknowns from three to one, and it also automatically removes artefacts
in finite-element methods (FEM) caused by the displacement formulation \citep{HaOuVe78}.
This is crucial since it is the main reason for which a displacement formulation cannot be used in fluid regions in a FEM.
It also permits acoustic-elastic coupling based on a non-iterative time scheme \citep{ChVa04}.
Displacement is then: u = grad(Chi) / rho,
velocity is then: v = grad(Chi\_dot) / rho (Chi\_dot being the time derivative of Chi)
and pressure is: p = - Chi\_dot\_dot  (Chi\_dot\_dot being the time second derivative of Chi).
First-order acoustic-acoustic discontinuities are also handled automatically
because pressure is continuous at such an interface, therefore Chi\_dot\_dot
is continuous, therefore Chi is also continuous, which is consistent with
the spectral-element basis functions and with the assembling process.
This is the reason why a simple displacement potential u = grad(Chi) would
not work because it would be discontinuous at such an interface and would
therefore not be consistent with the basis functions. And first-order acoustic-acoustic discontinuities are used
routinely in several domains, e.g. in the oil industry when dealing with geological models as acoustic only (no S waves),
or in ocean acoustics when considering the ocean bottom as fluid (again, no S waves).

@@

A solution to this could be to work with pressure as the scalar unknown (\cite{God11}, equation 23),
or even better with the time integral (the primitive) of pressure (see e.g. \cite{Eve81});
this was the approach that was originally implemented in SPECFEM until 10 years ago or so, but \cite{ChVa04} showed that switching to a displacement potential
or to a potential of $\rho \mathbf{u}$ enables one to couple the fluid and solid regions without iterating on the coupling condition in the time scheme,
thus resulting in much easier and cheaper implementation; since at that time SPECFEM users were not studying media with strong local variations of density (in solid Earth
geophysics such large local variations in fluid regions never happen), the SPECFEM developers permanently switched to such a displacement-based potential formulation.
However, for media with strong local variations of density in a fluid region we should reconsider the issue and consider trying the pressure-based formulation again
to see if there is a way of using it without having to numerically iterate on the fluid-solid matching condition.

Another possibility, suggested by Emmanuel Chaljub, would be to use two potentials instead of a single one (see Section~2.4 of \cite{ChKoViCaVaFe07}).
